---
title: "Project 1"
author: "Vatsal Dalal"
date: "2023-10-27"
output: html_document
---

```{r}
library(tidyverse)
library(dplyr)
library(nycflights13)
library(ggplot2)
library(resampledata3)
library(skimr)
```

## Step 1. Create data Frame for the Analysis:
```{r}
data("weather")
df <-merge(NycFlights, weather, by.NycFlights = c("origin", "time_hour"), by.weather = c("origin", "time_hour"), all.x = FALSE, all.y = FALSE, sort = TRUE)
Ua_df <- df %>%   filter(carrier=="UA")
# Remove rows with NA values
#Ua_df <- na.omit(Ua_df)
Ua_df
```
## Step 2 EDA

```{r}
ua_data <- Ua_df
# Filter the data fo
skim(ua_data)
```


```{r}
# EDA: Time of Day Analysis
library(dplyr)

# Categorize Time of Day
ua_data <- ua_data %>%
  mutate(time_of_day_category = case_when(
    hour(time_hour) >= 5 & hour(time_hour) < 12 ~ "Morning",
    hour(time_hour) >= 12 & hour(time_hour) < 17 ~ "Afternoon",
    hour(time_hour) >= 17 & hour(time_hour) < 21 ~ "Evening",
    TRUE ~ "Night"
  ))

# Categorize Time of Year (Season)
ua_data<- ua_data %>%
  mutate(
    month = as.numeric(month), # Convert the "month" variable to numeric if it's not already
    season = case_when(
      between(month, 1, 3) ~ "Winter",
      between(month, 4, 6) ~ "Spring",
      between(month, 7, 9) ~ "Summer",
      between(month, 10, 12) ~ "Fall"
    )
  )

# Categorize Temperature
ua_data <- ua_data %>%
  mutate(temperature_category = case_when(
    temp > 0 & temp <= 20 ~ "cold",
    temp > 20 ~ "Hot"
  ))

# Remove rows with null values in wind_speed column
ua_data <- ua_data %>%
  filter(!is.na(wind_speed)) %>%
  mutate(wind_speed_category = case_when(
    wind_speed <= 15 ~ "Calm",
    wind_speed > 15 ~ "Strong"
  ))



# Categorize Precipitation
ua_data <- ua_data %>%
  mutate(precipitation_category = case_when(
    precip == 0 ~ "No Precipitation",
    TRUE ~ "Precipitation"  # Everything else is categorized as "Precipitation"
  ))


# Categorize Visibility
ua_data <- ua_data %>%
  mutate(visibility_category = case_when(
    visib > 2 ~ "Good Visibility",  # Visibility greater than 2 is considered good
    TRUE ~ "Poor Visibility"  # Everything else is categorized as "Poor Visibility"
  ))


ua_data
```


```{r}
# Create a summary of time of day categories and departure delays
time_of_day_summary <- ua_data %>%
  group_by(time_of_day_category) %>%
  summarise(
    Total_Flights = n(),
    Mean_Dep_Delay = mean(dep_delay, na.rm = TRUE)
  )
time_of_day_summary
# Create a bar plot
ggplot(time_of_day_summary, aes(x = time_of_day_category, y = Mean_Dep_Delay)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Mean Departure Delay by Time of Day",
    x = "Time of Day Category",
    y = "Mean Departure Delay (minutes)"
  ) +
  theme_minimal()





```



```{r}

# Group by month and calculate the mean departure delay
monthly_delay_summary <- ua_data %>%
  group_by(month) %>%
  summarize(mean_delay = mean(dep_delay, na.rm = TRUE))

# Create a bar plot to visualize the mean departure delay by month
ggplot(monthly_delay_summary, aes(x = factor(month), y = mean_delay)) +
  geom_bar(stat = "identity", fill = "black") +
  xlab("Month") +
  ylab("Mean Departure Delay (minutes)") +
  ggtitle("Mean Departure Delay by Month")

```

```{r}
# Create a line plot for Season categories vs. Mean Departure Delay
ua_data %>%
  group_by(season) %>%
  summarize(mean_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = season, y = mean_dep_delay, group = 1)) +
  geom_line() +
  labs(title = "Mean Departure Delay by Season", x = "Season", y = "Mean Departure Delay")

```



```{r}
# Calculate the mean departure delay for each temperature
# Create a scatter plot for Temperature categories vs. Mean Departure Delay
ua_data %>%
  group_by(temperature_category) %>%
  reframe(mean_dep_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = temperature_category, y = mean_dep_delay)) +
  geom_point() +
  labs(title = "Mean Departure Delay by Temperature Category", x = "Temperature Category", y = "Mean Departure Delay")
```

```{r}
# Calculate the mean departure delay for each wind speed
# Create a bar plot for Wind Speed categories vs. Mean Departure Delay
ua_data %>%
  group_by(wind_speed_category) %>%
  summarize(mean_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = wind_speed_category, y = mean_dep_delay)) +
  geom_bar(stat = "identity") +
  labs(title = "Mean Departure Delay by Wind Speed Category", x = "Wind Speed Category", y = "Mean Departure Delay")
```

```{r}
# Create a bar plot for Precipitation categories vs. Mean Departure Delay
ua_data %>%
  group_by(precipitation_category) %>%
  summarize(mean_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = precipitation_category, y = mean_dep_delay)) +
  geom_bar(stat = "identity") +
  labs(title = "Mean Departure Delay by Precipitation Category", x = "Precipitation Category", y = "Mean Departure Delay")

```

```{r}
# Create a line plot for Visibility categories vs. Mean Departure Delay
ua_data %>%
  group_by(visibility_category) %>%
  summarize(mean_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = visibility_category, y = mean_dep_delay)) +
  geom_bar(stat = "identity") +
  labs(title = "Mean Departure Delay by Visibility Category", x = "Visibility Category", y = "Mean Departure Delay")
```

## Permutation Testing for Time of delay analysis


## Permutation Testing in between morning and evening:
```{r}
N<- 10^4-1 
observed <- mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Evening'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Morning'], na.rm = TRUE)
result <- numeric(N) 
for (i in 1:N) 
{   
  index <- sample(nrow(ua_data),
  size=nrow(ua_data %>% filter(time_of_day_category == 'Morning')), replace = FALSE)   
  result[i] <- mean(ua_data$dep_delay[index], na.rm = TRUE) - mean(ua_data$dep_delay[-index], na.rm = TRUE) 
}
ggplot(data = tibble(result), mapping = aes(x = result)) +  
geom_histogram() +   
geom_vline(xintercept = observed, color = "red")
2 * ((sum(result >= observed) + 1) / (N + 1))
```

## Permutation Testing in between morning and Afternoon
```{r}
N<- 10^4-1 
observed <- mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Afternoon'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Morning'], na.rm = TRUE)
result <- numeric(N) 
for (i in 1:N) 
{   
  index <- sample(nrow(ua_data),
  size=nrow(ua_data %>% filter(time_of_day_category == 'Morning')), replace = FALSE)   
  result[i] <- mean(ua_data$dep_delay[index], na.rm = TRUE) - mean(ua_data$dep_delay[-index], na.rm = TRUE) 
}
ggplot(data = tibble(result), mapping = aes(x = result)) +  
geom_histogram() +   
geom_vline(xintercept = observed, color = "red")
2 * ((sum(result >= observed) + 1) / (N + 1))
```


## Permutation Testing in between Morning and Night
```{r}
N<- 10^4-1 
observed <- mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Night'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Morning'], na.rm = TRUE)
result <- numeric(N) 
for (i in 1:N) 
{   
  index <- sample(nrow(ua_data),
  size=nrow(ua_data %>% filter(time_of_day_category == 'Morning')), replace = FALSE)   
  result[i] <- mean(ua_data$dep_delay[index], na.rm = TRUE) - mean(ua_data$dep_delay[-index], na.rm = TRUE) 
}
ggplot(data = tibble(result), mapping = aes(x = result)) +  
geom_histogram() +   
geom_vline(xintercept = observed, color = "red")
2 * ((sum(result >= observed) + 1) / (N + 1))
```
## Permutation Testing in between Afternoon and Evening
```{r}
N<- 10^4-1 
observed <- mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Evening'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Afternoon'], na.rm = TRUE)
result <- numeric(N) 
for (i in 1:N) 
{   
  index <- sample(nrow(ua_data),
  size=nrow(ua_data %>% filter(time_of_day_category == 'Afternoon')), replace = FALSE)   
  result[i] <- mean(ua_data$dep_delay[index], na.rm = TRUE) - mean(ua_data$dep_delay[-index], na.rm = TRUE) 
}
ggplot(data = tibble(result), mapping = aes(x = result)) +  
geom_histogram() +   
geom_vline(xintercept = observed, color = "red")
2 * ((sum(result >= observed) + 1) / (N + 1))
```

## Permutation Testing in between Afternoon and Night
```{r}
N<- 10^4-1 
observed <- mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Night'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Afternoon'], na.rm = TRUE)
result <- numeric(N) 
for (i in 1:N) 
{   
  index <- sample(nrow(ua_data),
  size=nrow(ua_data %>% filter(time_of_day_category == 'Afternoon')), replace = FALSE)   
  result[i] <- mean(ua_data$dep_delay[index], na.rm = TRUE) - mean(ua_data$dep_delay[-index], na.rm = TRUE) 
}
ggplot(data = tibble(result), mapping = aes(x = result)) +  
geom_histogram() +   
geom_vline(xintercept = observed, color = "red")
2 * ((sum(result >= observed) + 1) / (N + 1))
```


## Permutation Testing in between Afternoon and Night
```{r}
N<- 10^4-1 
observed <- mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Night'], na.rm = TRUE)- mean(ua_data$dep_delay[ua_data$time_of_day_category == 'Afternoon'], na.rm = TRUE)
result <- numeric(N) 
for (i in 1:N) 
{   
  index <- sample(nrow(ua_data),
  size=nrow(ua_data %>% filter(time_of_day_category == 'Afternoon')), replace = FALSE)   
  result[i] <- mean(ua_data$dep_delay[index], na.rm = TRUE) - mean(ua_data$dep_delay[-index], na.rm = TRUE) 
}
ggplot(data = tibble(result), mapping = aes(x = result)) +  
geom_histogram() +   
geom_vline(xintercept = observed, color = "red")
2 * ((sum(result >= observed) + 1) / (N + 1))
```




## Permutation Testing for Time of year analysis

## Permutation Testing in between winter and spring
```{r}
N<- 10^4-1 
observed1 <- mean(ua_data$dep_delay[ua_data$season == 'Spring'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$season == 'Winter'], na.rm = TRUE)
result1 <- numeric(N) 
for (i in 1:N) 
{   
  index <- sample(nrow(ua_data),
  size=nrow(ua_data %>% filter(season == 'Winter')), replace = FALSE)   
  result1[i] <- mean(ua_data$dep_delay[index], na.rm = TRUE) - mean(ua_data$dep_delay[-index], na.rm = TRUE) 
}
ggplot(data = tibble(result1), mapping = aes(x = result1)) +  
geom_histogram() +   
geom_vline(xintercept = observed1, color = "red")
2 * ((sum(result1 >= observed1) + 1) / (N + 1))
```

## Permutation Testing in between Winter and summer
```{r}
N<- 10^4-1 
observed2 <- mean(ua_data$dep_delay[ua_data$season == 'Winter'], na.rm = TRUE) - mean(ua_data$dep_delay[ua_data$season == 'Summer'], na.rm = TRUE)
result2 <- numeric(N) 
for (i in 1:N) 
{   
  index <- sample(nrow(ua_data),
  size=nrow(ua_data %>% filter(season == 'Winter')), replace = FALSE)   
  result2[i] <- mean(ua_data$dep_delay[index], na.rm = TRUE) - mean(ua_data$dep_delay[-index], na.rm = TRUE) 
}
ggplot(data = tibble(result2), mapping = aes(x = result2)) +  
geom_histogram() +   
geom_vline(xintercept = observed2, color = "red")
2 * ((sum(result2 <= observed2) + 1) / (N + 1))
```
## Permutation Testing in between Winter and Fall
```{r}
N<- 10^4-1 
observed3 <- mean(ua_data$dep_delay[ua_data$season == 'Winter'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$season == 'Fall'], na.rm = TRUE)
result3 <- numeric(N) 
for (i in 1:N) 
{   
  index1 <- sample(nrow(ua_data),
  size =nrow(ua_data %>% filter(season == 'Winter')), replace = FALSE)   
  result3[i] <- mean(ua_data$dep_delay[index1], na.rm = TRUE) - mean(ua_data$dep_delay[-index1], na.rm = TRUE) 
}
ggplot(data = tibble(result3), mapping = aes(x = result3)) +  
geom_histogram() +   
geom_vline(xintercept = observed3, color = "red")
2 * ((sum(result3 <= observed3) + 1) / (N + 1))
```


## Permutation Testing in between Spring and summer 
```{r}
N<- 10^4-1 
observed4 <- mean(ua_data$dep_delay[ua_data$season == 'Spring'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$season == 'Summer'], na.rm = TRUE)
result4 <- numeric(N) 
for (i in 1:N) 
{   
  index2 <- sample(nrow(ua_data),
  size =nrow(ua_data %>% filter(season == 'Spring')), replace = FALSE)   
  result4[i] <- mean(ua_data$dep_delay[index2], na.rm = TRUE) - mean(ua_data$dep_delay[-index2], na.rm = TRUE) 
}
ggplot(data = tibble(result4), mapping = aes(x = result3)) +  
geom_histogram() +   
geom_vline(xintercept = observed4, color = "red")
2 * ((sum(result4 >= observed4) + 1) / (N + 1))
```

## Permutation Testing in between Spring and Fall
```{r}
N<- 10^4-1 
observed5 <- mean(ua_data$dep_delay[ua_data$season == 'Spring'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$season == 'Fall'], na.rm = TRUE)
result5 <- numeric(N) 
for (i in 1:N) 
{   
  index3 <- sample(nrow(ua_data),
  size =nrow(ua_data %>% filter(season == 'Spring')), replace = FALSE)   
  result5[i] <- mean(ua_data$dep_delay[index1], na.rm = TRUE) - mean(ua_data$dep_delay[-index1], na.rm = TRUE) 
}
ggplot(data = tibble(result5), mapping = aes(x = result5)) +  
geom_histogram() +   
geom_vline(xintercept = observed5, color = "red")
2 * ((sum(result5 >= observed5) + 1) / (N + 1))
```


## Permutation Testing in between summer and Fall
```{r}
N<- 10^4-1 
observed6 <- mean(ua_data$dep_delay[ua_data$season == 'Summer'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$season == 'Fall'], na.rm = TRUE)
result6 <- numeric(N) 
for (i in 1:N) 
{   
  index4 <- sample(nrow(ua_data),
  size =nrow(ua_data %>% filter(season == 'Summer')), replace = FALSE)   
  result6[i] <- mean(ua_data$dep_delay[index4], na.rm = TRUE) - mean(ua_data$dep_delay[-index4], na.rm = TRUE) 
}
ggplot(data = tibble(result6), mapping = aes(x = result6)) +  
geom_histogram() +   
geom_vline(xintercept = observed6, color = "red")
2 * ((sum(result6 >= observed6) + 1) / (N + 1))
```

## Permutation testing on temprature Hot and warm:
```{r}
ua_data <- ua_data %>%
  filter(!is.na(temperature_category))
N<- 10^4-1 
observed7 <- mean(ua_data$dep_delay[ua_data$temperature_category == 'Hot'], na.rm = TRUE) -  mean(ua_data$dep_delay[ua_data$temperature_category == 'Cold'], na.rm = TRUE)
observed7
result7 <- numeric(N) 
for (i in 1:N) 
{   
  index5 <- sample(nrow(ua_data),
  size =nrow(ua_data %>% filter(temperature_category == 'Cold')), replace = FALSE)   
  result7[i] <- mean(ua_data$dep_delay[index5], na.rm = TRUE) - mean(ua_data$dep_delay[-index5], na.rm = TRUE) 
}
ggplot(data = tibble(result7), mapping = aes(x = result7)) +  
geom_histogram() +   
geom_vline(xintercept = observed7, color = "red")
2*((sum(result7 >= observed7) + 1) / (N + 1))
```



## Permutation test for wind_speed
```{r}
N<- 10^4-1 
observed8 <- mean(ua_data$dep_delay[ua_data$wind_speed_category == 'Calm'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$wind_speed_category == 'Strong'], na.rm = TRUE)
result8 <- numeric(N) 
for (i in 1:N) 
{   
  index6 <- sample(nrow(ua_data),
  size =nrow(ua_data %>% filter(wind_speed_category == 'Calm')), replace = FALSE)   
  result8[i] <- mean(ua_data$dep_delay[index6], na.rm = TRUE) - mean(ua_data$dep_delay[-index6], na.rm = TRUE) 
}
ggplot(data = tibble(result8), mapping = aes(x = result8)) +  
geom_histogram() +   
geom_vline(xintercept = observed8, color = "red")
2*((sum(result8 <= observed8) + 1) / (N + 1))
```

## Permutation test for percip
```{r}
N<- 10^4-1 
observed9 <- mean(ua_data$dep_delay[ua_data$precipitation_category == 'No Precipitation'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$precipitation_category == 'Precipitation'], na.rm = TRUE)
result9 <- numeric(N) 
for (i in 1:N) 
{   
  index7 <- sample(nrow(ua_data),
  size =nrow(ua_data %>% filter(precipitation_category == 'Precipitation')), replace = FALSE)   
  result7[i] <- mean(ua_data$dep_delay[index7], na.rm = TRUE) - mean(ua_data$dep_delay[-index7], na.rm = TRUE) 
}
ggplot(data = tibble(result9), mapping = aes(x = result9)) +  
geom_histogram() +   
geom_vline(xintercept = observed9, color = "red")
2*((sum(result9 <= observed9) + 1) / (N + 1))
```

## Permutation test for visibility
```{r}
N<- 10^4-1 
observed10 <- mean(ua_data$dep_delay[ua_data$visibility_category == 'Good Visibility'], na.rm = TRUE)-mean(ua_data$dep_delay[ua_data$visibility_category == 'Poor Visibility'], na.rm = TRUE)
result10 <- numeric(N) 
for (i in 1:N) 
{   
  index8 <- sample(nrow(ua_data),
  size =nrow(ua_data %>% filter(visibility_category == 'Good Visibility')), replace = FALSE)   
  result10[i] <- mean(ua_data$dep_delay[index8], na.rm = TRUE) - mean(ua_data$dep_delay[-index8], na.rm = TRUE) 
}
ggplot(data = tibble(result10), mapping = aes(x = result10)) +  
geom_histogram() +   
geom_vline(xintercept = observed10, color = "red")
2*((sum(result10 <= observed10) + 1) / (N + 1))
```

